Contracts for unions and intersections are hard for Nickel

### 2021-06-04 paper.tex
Continue reworking of section 4 [WIP]

Static typing is about enforcing that some property always holds for a
potentially unknown value, no matter the execution part. A dynamic check, on the
other hand, just need to verify a property for a specific and known value, which
is usually much easier. For example, dependent function types are complex and
mostly implemented in proof assistants (although they are slowly making their
way in some mainstream languages), whereas the corresponding dependent function
contracts can be implemented in a few line of code (as is done in
Racket~\cite{}).

@@

\subsection*{Union contracts as an effect}

In a pure language with standard contracts, the failure of a function contract
can always be blamed on a single call. Figure~\ref{fig:pos-to-pos} shows a
function with a simple contract attached that ensures that both the argument and
the result of \nickel{f} are positive numbers. \nickel{f} is called twice, and
the second call violates the requirement on the codomain, leading to blame
\nickel{f}. The first call to \nickel{f} does not matter, and \nickel{5} is a
single and independent witness of the violation of the contract. The user can be
pointed to one location in practice.

There is an intuitive justification for this fact. Apart from the error
reporting part (but this is the crucial in practice!), the current contract
system of Nickel can be implemented purely as a library, just given a
\nickel{fail} primitive to abort the execution. In particular, applying a
function contract to \nickel{f} replaces it with a guarding function
\nickel{fWrapped} that performs the required checks. Thus, in a pure setting,
the failure of \nickel{fWrapped 5} must be independant of its environment, and
in particular of the call \nickel{fWrapped 10}.

@@

Union contracts are a different matter. Consider the program presented in
Figure \ref{fig:wrong-union-function}. \nickel{f} is again violating the
contract, as it neither maps positive numbers to positive numbers nor positive
numbers to negative numbers.

@@

In a system with unions, this contract would fail, because \nickel{f 10}
(returns \nickel{3}) is a witness of \nickel{f} failing the contract
\nickel{Positive -> NonPositive}, and \nickel{f 5} (returns \nickel{-2}) is a
witness of \nickel{f} failing \nickel{Positive -> Positive}.  But, as opposed to
the previous example, removing one call or the other will make the program pass!
Indeed, the first call only fails the \nickel{Positive -> NonPositive} part of
the union, while the second one only fails the \nickel{Positive -> NonPositive}
part. Both are needed to prove the violation of the full contract.

Even in a pure setting, unions introduce a form of effect at a distance: the
result of \nickel{f 5} is not local anymore, as it now depends on the previous
execution of other calls to \nickel{f}. Unions contracts break the
equation~\ref{eq:cbn-expansion} of Section~\ref{sec:optimizations}, and the
candidate example of Figure~\ref{fig:optimizations-let-floating-ex} cannot be
optimized in general.

Figure~\ref{fig:optimized-programs} illustrates this point further. It comprises
an original program, and one where the common subexpression \nickel{f 1} have
been eliminated. These two programs behave very differently:
\begin{itemize}
    \item The original version returns \nickel{(1, "False")} without failing.
    \item The optimized version fails with a contract violation.
\end{itemize}

since the single application of \nickel{f 1}, applied to both \nickel{true} and
\nickel{false} will be able to find the error on the union contract
\nickel{(Bool -> Num) @| (Bool -> Str)}.

@@

@@

To sum up, union contracts introduce a form of effect at a distance. This
impacts both the programmer and the compiler or interpreter:
\begin{itemize}
    \item As for any side-effect, this breaks local reasoning.
    \item It prevents optimizations.
\end{itemize}
\todo{WIP}

A valid argument against this point would be stating that further work should be done
on optimizers, so to not modify these tricky cases, even if this is true,
programmers also modify their code on a daily basis, and if moving from the first
code snippet at Figure \ref{fig:optimized-programs} to the second one changes
the behaviour of a program, we argue that union contracts not only go against the
implementation of languages, but also against their usability.

Summing it up, wrapping a value with a union of higher order contracts means that
that value now has a state, that may get modified silently at every use site.
Both reviewed papers suffer from this defect since it is the semantics
of unions that interact badly with code manipulation operations,
not a particular implementation.
This addition of global state complicates both the work of the language
implementor, as well as the language user, since it adds non trivial
side effects to your code.

\subsection*{User defined contracts do not intersect}
\label{sec:flat-and-inter}

Stop for a second and think how you could implement an intersection
contract. The first thought you may have is "just apply
both contracts", as shown on the Implementation snippet
from Figure \ref{fig:inter-contracts}.
Of course, if it were that simple we wouldn't be writing this paper.
This idea starts to show problems when you first combine it with
higher order contracts, since it would mean not complying with
at least one of these contracts negatively would entail breaking the
whole thing, as shown on the second example on the same Figure,
that would fail.

>>>>

Static typing is about enforcing that some property about a (potentially
unknown) value always holds, no matter the execution path. A dynamic check, on
the other hand, just need to verify a property for a specific and known value,
which is usually much easier. For example, dependent function types are complex
and mostly implemented in proof assistants (although they are slowly making
their way in some mainstream languages), whereas the corresponding dependent
function contracts can be implemented in a few line of code (they are
implemented in e.g. Racket~\cite{}).

@@

\subsection*{Union contracts as a side-effect}

In Nickel, the failure of a function contract can always be traced back to a
single call. For example, take the function \nickel{f} with a simple contract
attached of Figure~\ref{fig:pos-to-pos}. The program fails with a contract error
blaming \nickel{f}. Indeed, the second call \nickel{f 5} fails because the
return value violates the \nickel{Positive} contract. The first call to
\nickel{f} does not matter, and \nickel{f 5} is a single and independent witness
of the violation of the contract. The user can be pointed to one location in
practice.

This single witness property can be justified as follows. Apart from the error
reporting part (although this is the crucial bit in practice!), the current
contract system of Nickel can be implemented purely as a library, requiring only
a \nickel{fail} primitive to abort the execution. In practice, applying a
function contract to \nickel{f} replaces it with a \nickel{f'} that additionally
performs the required checks. Thus, since the core language is pure (the
presence of \nickel{fail} isn't relevant here), the failure of \nickel{f' 5}
must be independent of its environment and of any previous call to \nickel{f'}.

@@

Union contracts are different. Consider the program presented in
Figure~\ref{fig:wrong-union-function}. The same \nickel{f} is now given a union
contract. \nickel{f} is violating this contract once again, as it neither maps
positive numbers to positive numbers nor positive numbers to negative numbers.

@@

This program must fail, because \nickel{f 10} is a witness of \nickel{f} failing
the contract \nickel{Positive -> NonPositive}, and \nickel{f 5} is a witness of
\nickel{f} failing \nickel{Positive -> Positive}.  But, as opposed to
Figure~\ref{fig:pos-to-pos}, removing only one the call makes the program
succeed! Indeed, the first call only fails the \nickel{Positive -> NonPositive}
part of the union, while the second one only fails the \nickel{Positive ->
NonPositive} part. In this particular example, there can't be a single call that
is a witness of the violation of the contract: a minimum of two are needed.

This behavior indicates that union contracts introduce side-effects. The result
of \nickel{f 5} now depends on the previous execution and more generally on any
prior call to \nickel{f}. This behavior of union contracts break the
equation~\ref{eq:cbn-expansion} of Section~\ref{sec:optimizations}. The
candidate example of Figure~\ref{fig:optimizations-let-floating-ex} can't be
optimized in general.

Figure~\ref{fig:optimized-programs} illustrates this point further. There is an
original program, and an optimized version the common subexpression \nickel{f 1}
has been eliminated. While equivalent in a pure language without contracts,
these two programs behave differently here:

\begin{itemize}
    \item The original version returns \nickel{(1, "False")} without failing.
    \item The optimized version fails with a contract violation.
\end{itemize}

In the original version, each partial application \nickel{f 1} gives rise to a
fresh instance of the contract \nickel{Bool -> Num @| Bool -> Str}. These
instances are independent, and can pick a different component of the union to
satisfy. Although \nickel{f} doesn't actually respect the contract, these calls
are not enough to prove so. In the optimized version, \nickel{g} is endowed with
a single contract, that must pick one of the two components of the union. There,
the two calls refers to the same union contracts, and shows that \nickel{f}
violate its initial contract.

@@

To sum up, the addition of union contracts introduce side-effects in a pure
language. Side-effects have well-known pitfalls: 
\begin{itemize}
    \item For the programmer, they are hard to reason about. They prevent local
        reasoning. In our previous examples, removing or adding a function call
        somewhere can toggle a failure in a call at a totally different
        location.
    \item For the interpreter (or compiler), they inhibit many optimizations and
        program transformations.
\end{itemize}

% A valid argument against this point would be stating that further work should be done
% on optimizers, so to not modify these tricky cases, even if this is true,
% programmers also modify their code on a daily basis, and if moving from the first
% code snippet at Figure \ref{fig:optimized-programs} to the second one changes
% the behaviour of a program, we argue that union contracts not only go against the
% implementation of languages, but also against their usability.
%
% Summing it up, wrapping a value with a union of higher order contracts means that
% that value now has a state, that may get modified silently at every use site.
% Both reviewed papers suffer from this defect since it is the semantics
% of unions that interact badly with code manipulation operations,
% not a particular implementation.
% This addition of global state complicates both the work of the language
% implementor, as well as the language user, since it adds non trivial
% side effects to your code.

\subsection*{Intersection of user-defined contracts}
\label{sec:flat-and-inter}

A natural - but naive - implementation of intersection contracts could be the
following: to apply a contract \nickel{A @& B}, apply both contracts \nickel{A}
and \nickel{B} sequentially. This intuition works for simple contracts: surely
checking that \nickel{x | Natural @& Odd} amounts to check that \nickel{x |
Natural} and \nickel{x | Odd}. Unfortunately, this doesn't scale to higher-order
contracts.

@@

works for a lot of simple contracts: Stop for
a second and think how you could implement an intersection contract. The first
thought you may have is "just apply both contracts", as shown on the
Implementation snippet from Figure \ref{fig:inter-contracts}.  Of course, if it
were that simple we wouldn't be writing this paper.  This idea starts to show
problems when you first combine it with higher order contracts, since it would
mean not complying with at least one of these contracts negatively would entail
breaking the whole thing, as shown on the second example on the same Figure,
that would fail.
